---
title: "Other notes from Bioinformatics Summer School 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Day 1 

## Morgan, intro R


`methods` lists all methods available for a given class, but not simple functions:

```{r}
methods(class = lm)
```
Getting specific help for a method call:
```{r, eval = FALSE}
?plot.lm
```

Note that not all functions are shown e.g. abline, which is a pure function:

```{r}
abline
```
where e.g. `anova` is a generic that dispatches to various methods:
```{r}
anova
```

and methods is actually listing the generics that operate. 
 
## Bioconductor intro

`nchar` counts the number of characters in each entry of character vector:

```{r}
seq = c("AAACA", "CATGC")
nchar(seq)
```

```{r}
"DNAStringSet"
```

Check available packages (not case sensitive for search):

```{r}
library(BiocManager)
BiocManager::available("readr")
```


## classes, methods etc



```{r}
class(anova); anova
```

anova is an S3 class (all things such as generic.method are S3)

```{r}
fit<- lm(mpg ~ cyl, cars)
```


```{r}
getAnywhere(anova.lm)
```
```{r}
library(Biostrings)
?`reverseComplement,DNAStringSet-class`  
```
This structure is for S4:
```{r}
?`reverseComplement,DNAStringSet-method`
```

